GeckoGraph: A Visual Language for Polymorphic Types

### 2025-04-06 main.tex
Citations

Although polymorphic typing promises robustness and a high degree of code reusability, studies~\cite{Jun2000-ec, Jun2000-yu} show that using polymorphism in practice poses challenges, especially for novice users. These studies have shown that humans tend to focus on concrete types and only rely on polymorphic type checking as a last resort. In practice, polymorphic types often pose usability problems for programmers. New polymorphic type variables can be created during type checking. These intermediate type variables are often kept behind the curtain unless a type error is encountered. This often results in programmers resolving type errors with type variables not authored by the programmers themselves (Fig. \ref{fig:example-foldable}).

@@

While type polymorphism is one of the oldest topics in programming language theory \cite{Cardelli1987-fp}, little research focuses on the \textit{usability} of polymorphic types. Hage argues that the expressiveness and power of type systems often come at the cost of usability~\cite{Hage2020-hg}. We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern HCI techniques. To achieve this, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. Our study evaluating 714 participants' ability to solve a series of programming challenges using GeckoGraph versus text-based type annotation is, to our knowledge, the largest controlled user study of a functional programming tool or concept. The study results show that GeckoGraph helps improve programmers' ability to succeed in resolving the type challenges, especially for novice programmers.

@@

\subsection{Design of GeckoGraph}
The design of GeckoGraph focuses on visualizing types in functional languages (e.g., Haskell, ML). In this paper, we use Haskell as an example. As illustrated in this section, it can express basic types, polymorphic types, algebraic data types, and many advanced type-level features. However, GeckoGraph could also be used in imperative and multiparadigm languages such as TypeScript and Rust. We provide a GeckoGraph visualization tool avaiable online for multiple languages \todo{cite GeckoGraph website}.

@@

The experiment uses a between-subject design. However, all participants receive both treatments (with and without GeckoGraph) during their runs. Participants are assigned to one of two groups. Both groups receive the same tasks in the same order. Group one participants are assisted by GeckoGraph on even levels and only text-based type annotation on odd levels. Group two participants are the same but with the order flipped. Both groups have access to the text-based type signature for all tasks. The number of participants in the two groups is counterbalanced.

\subsection{Hypothesis}
In programming tasks that involve reading and understanding polymorphic types, graphic notation using visual elements that provide higher grouping strength (colors, shapes, sizes, and symbols) can improve the performance of such tasks compared to traditional text-based type notation. Our null hypothesis is: {\it Using graphic notation has no effect compared to traditional text-based type notation.} This hypothesis and the task design were registered at the Open Science Foundation prior to data collection \todo{cite this}.

\subsection{Task Design} \label{subsection:task}
Participants in both groups receive the same 10 tasks. The tasks start off easy but gradually increase in difficulty.  In each task, a target type signature of the function \texttt{zeroToHero} is given to the participants. Participants are provided with a list of available functions to implement the target function. This is to simulate the tasks of selecting useful functions from a library. In addition, participants are not allowed to use any other functions or variables outside the provided functions; even the Haskell prelude is not available. This ensures that everyone has the same knowledge and minimizes the effect of familiarity.

@@

Data collection from the human study was stopped after the planned cut-off period of 14 days. After the cut-off date, the ZeroToHero game is open source and available for free evaluation \cite{Anonymous_undated-ne} and repeating our experiment, but no further data was collected.

>>>>

While promoting robustness and code reusability, polymorphic types present challenges in comprehension, particularly for novice users \cite{Jun2000-ec, Jun2000-yu}. Research indicates that humans tend to focus on concrete types and rely on polymorphic type checking only as a last resort. Moreover, polymorphic types often complicate type error usability. During type checking, compilers frequently generate new polymorphic type variables temporarily. These variables are typically discarded after type checking unless a type error occurs, in which case they are referenced in error messages. Consequently, programmers often face the task of resolving type errors involving type variables they did not explicitly author (Fig. \ref{fig:example-foldable}).

@@

While type polymorphism is one of the oldest topics in programming language theory \cite{Cardelli1987-fp}, little research focuses on the \textit{usability} of polymorphic types. It is often argued that the expressiveness and power of type systems often come at the cost of usability~\cite{Hage2020-hg}. We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern HCI techniques. In this paper, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. Our study evaluating 714 participants' ability to solve a series of programming challenges using GeckoGraph versus text-based type annotation is, to our knowledge, the largest controlled user study of a functional programming tool or concept. The study results show that GeckoGraph helps improve programmers' ability to succeed in resolving the type challenges, especially for novice programmers.

@@

\subsection{Design of GeckoGraph}
The design of GeckoGraph focuses on visualizing types in functional languages (e.g., Haskell, ML). In this paper, we use Haskell as an example. As illustrated in this section, it can express basic types, polymorphic types, algebraic data types, and many advanced type-level features. However, GeckoGraph could also be used in imperative and multiparadigm languages such as TypeScript and Rust. We provide a GeckoGraph visualization tool avaiable online for multiple languages \cite{Fu2025-ka}.

@@

The experiment uses a between-subject design. Participants are assigned to one of two groups. The tasks asiigned to the two groups are the same order. Participants receive both treatments (with and without GeckoGraph) during their runs but in different order.   Group one is assisted by GeckoGraph on even levels and text-only type annotation on odd levels. Group two participants are the same but with the order flipped. The number of participants in the two groups is counterbalanced.

\subsection{Hypothesis}
In programming tasks that involve reading and understanding polymorphic types, graphic notation using visual elements that provide higher grouping strength (colors, shapes, sizes, and symbols) can improve the performance of such tasks compared to traditional text-based type notation. Our null hypothesis is: {\it Using graphic notation has no effect compared to traditional text-based type notation.} This hypothesis and the task design were registered at the Open Science Foundation prior to data collection \cite{Stuckey2023-vl}.

\subsection{Task Design} \label{subsection:task}
During the study, participants receive 10 tasks, represented by the 10 levels of the game {\it zeroToHero}. The tasks start off easy but gradually increase in difficulty.  In each task, a target type of the function \texttt{zeroToHero} is shown to the participants. Participants are provided with a list of available functions to implement the target function, and to ensure it assume the target type. This is to simulate the tasks of selecting useful functions from a library. Participants are not allowed to use any other functions or variables outside the provided functions; even the Haskell prelude is not available. This ensures that everyone has the same knowledge and minimizes the effect of familiarity.

@@

Data collection from the study was stopped after the planned cut-off period of 14 days. After the cut-off date, the ZeroToHero game is open source and available for free evaluation \cite{Anonymous_undated-ne} and repeating our experiment, but no further data was collected.
